\documentclass{scrartcl}

\usepackage[utf8]{inputenc}

\usepackage{natbib}
\usepackage{hyperref}
\usepackage{csquotes}
\usepackage{listings}
\usepackage{graphicx}
\usepackage[colorinlistoftodos]{todonotes}
% \usepackage{parskip}
% \setlength{\parskip}{10pt}
\usepackage{tikz}
\usetikzlibrary{arrows, decorations.markings}
\usepackage{chngcntr}
\counterwithout{figure}{section}


\begin{document}

\begin{titlepage}

\centering
{\scshape\LARGE Master thesis project proposal}

\vspace{0.5cm}
{\huge\bfseries Source-to-source translation between Agda and Idris
  }

\vspace{2cm}
{\Large Jakob Larsson \texttt{<jakob@karljakoblarsson.com>}}

\vspace{1.0cm}
{\large Suggested Supervisor at CSE: Patrik Jansson }

% \vspace{1.5cm}
\vspace{1.5cm}
{\large Relevant completed courses:}

{\itshape
Types for Programs and Proofs, DAT350 \\
Advanced Functional programming, TDA342 \\
}

\vfill
{\large \today}

\end{titlepage}

% \section{Introduction}

Agda~\cite{agda} and Idris~\cite{idris} are two dependently typed programming
languages.  Agda is mainly focused on automated theorem proving while Idris
prioritizes general purpose programming.  However, while both languages have
different focus, they have most of their features in common. The type systems
are very similar.

We wish to investigate if it is possible to construct a source to source
translation of a common subset from Idris to Agda.  A further goal is to do
bidirectional translation between the languages, but this is probably too
ambitious for this project.  The first challenge is to find a common subset for
which it is possible to translate. The task is to construct a compiler which
handles as much of this subset as possible.

% Idris to Agda as the main step. Bidirectional is probably hard.

% Målet är att göra översättningen åt båda hållen, fast det kanske är
% rimligt att fokusera åt ett håll från början. Det är såklart väldigt
% svårt, kanske omöjligt. Men det är svårt att svara på idag hur svårt
% det är.

This would allow Idris libraries to be reused in Agda. A Idris program may be
more easily proved in Agda. With bidirectional translation it would be possible
to use Idris to interface a well-proven Agda program with the world.  It will
also increase the confidence in a given proof if it is valid in both languages.

A small start is to just transpile small functions with non-dependent types
over Nat for example. Then we develop the implementation to support bigger
types. The goal then is to translate dependent types.  Functionality which is
not available in both languages should be represented by holes. That way the
user can implement whats missing.

To test the implementation we will use the Sequential Decision Problem library
IdrisLibs~\footnote{https://gitlab.pik-potsdam.de/botta/IdrisLibs} which is one
of the bigger Idris codebases available. This will guide our implementation by
first supporting the features used in the core parts of IdrisLibs. And then
work to support more and more of the library.

If possible we wish to reuse the Idris complier front-end for our
implementation, but there is always a challenge to interface with existing
code.  Another hard task is to find the biggest common subset of the languages.
It is trivial to name a few common features, bu a useful compiler needs to
support most widely used features. But there are incompatible differences
between Agda and Idris which complicates things.

% ~\cite{sdpcontrib}



% \section{Goals and Challenges}

% \section{Approach}



\bibliographystyle{plain}

\bibliography{ita}

\end{document}
